---
title: "Overlap at peak timepoints"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE)
library(tidyverse)
library(VennDiagram)
```

## Purpose

When you look at the number of genes changing in each timepoint, there is a striking pattern. In the WT cells regardless of Firre there is a bunch of genes change between 180 and 150 minutes. In the Firre KO cells regardless of Firre TG, there are a bunch of genes changing at 90 vs 60 minutes. I would like to know if these are the same genes that are changing in both the Firre induced and control cells. Then I would like to 
know if there is a common transcription factor driving these changes.

```{r}
dat <- read.csv("delta_de_shrunkenlfc_results.csv")
dat_sig <- dat %>% filter(padj < 0.1)
table(dat_sig$contrast)
```


```{r}
# Let's calculate the overlaps between cell types
dat_sig$timepoint <- gsub("_vs_", ";", gsub("timepoint_", ";",dat_sig$contrast))
dat_sig$timepoint <- sapply(dat_sig$timepoint, function(x) {
  unlist(strsplit(x, ";"))[[2]]
})
dat_sig <- dat_sig %>% select(contrast, timepoint, log2FoldChange, padj, gene_id, gene_name)

dat_sig$condition <- gsub("results/", ";", gsub("_shrunken", ";",dat_sig$contrast))
dat_sig$condition <- sapply(dat_sig$condition, function(x) {
  unlist(strsplit(x, ";"))[[2]]
})


# Okay, let's first check the KO 90vs60 timepoint.
table(dat_sig$condition)
ko_dox_90 <- dat_sig[which(dat_sig$timepoint == "90" & dat_sig$condition == "ko_dox"),"gene_id"]
ko_firre_90 <- dat_sig[which(dat_sig$timepoint == "90" & dat_sig$condition == "ko_firre"),"gene_id"]


v <- venn.diagram(list("ko_dox_90" = ko_dox_90, "ko_firre_90" = ko_firre_90), 
                       filename = NULL)
grid.newpage()
grid.draw(v)

# Hmm, are these 29 genes going in the same direction
sg <- intersect(ko_dox_90, ko_firre_90)
# That
sgdf <- dat_sig %>% filter(gene_id %in% sg, timepoint == "90") %>%
  select(condition, log2FoldChange, gene_name) %>%
  as.data.frame()

sgdf <- sgdf %>%
  spread(condition, log2FoldChange)
# Well damn. I'd say they line up pretty well. 
g <- ggplot(sgdf, aes(x = ko_dox, y = ko_firre))
g + geom_point()

```


```{r}
# Now let's check out the KO 180 timepoint
ko_dox_180 <- dat_sig[which(dat_sig$timepoint == "180" & dat_sig$condition == "ko_dox"),"gene_id"]
ko_firre_180 <- dat_sig[which(dat_sig$timepoint == "180" & dat_sig$condition == "ko_firre"),"gene_id"]

v <- venn.diagram(list("ko_dox_180" = ko_dox_180, "ko_firre_180" = ko_firre_180), 
                       filename = NULL)
grid.newpage()
grid.draw(v)

```

```{r}
# How about the WT 180 timepoint
wt_dox_180 <- dat_sig[which(dat_sig$timepoint == "180" & dat_sig$condition == "wt_dox"),"gene_id"]
wt_firre_180 <- dat_sig[which(dat_sig$timepoint == "180" & dat_sig$condition == "wt_firre"),"gene_id"]

v <- venn.diagram(list("wt_dox_180" = wt_dox_180, "wt_firre_180" = wt_firre_180), 
                       filename = NULL)
grid.newpage()
grid.draw(v)
# Hmm, are these 20 genes going in the same direction
sg <- intersect(wt_dox_180, wt_firre_180)
# That
sgdf <- dat_sig %>% filter(gene_id %in% sg, timepoint == "180") %>%
  select(condition, log2FoldChange, gene_name) %>%
  as.data.frame()

sgdf <- sgdf %>%
  spread(condition, log2FoldChange)
# Well damn. I'd say they line up pretty well. 
g <- ggplot(sgdf, aes(x = wt_dox, y = wt_firre))
g + geom_point() + geom_abline(slope = 1)

```

```{r}
sg_ko_90 <- intersect(ko_dox_90, ko_firre_90)
sg_wt_180 <- intersect(wt_dox_180, wt_firre_180)

v <- venn.diagram(list("sg_ko_90" = sg_ko_90, "sg_wt_180" = sg_wt_180), 
                       filename = NULL)
grid.newpage()
grid.draw(v)
intersect(sg_ko_90, sg_wt_180)
```

```{r}
# Hmm, what if I take the overlap of all the 90 min KO genes and all the 180 minute WT genes
wt_180 <- unique(dat_sig[which(dat_sig$timepoint == "180" & (dat_sig$condition == "wt_dox" | dat_sig$condition == "wt_firre")),"gene_id"])
ko_90 <- unique(dat_sig[which(dat_sig$timepoint == "90" & (dat_sig$condition == "ko_dox" | dat_sig$condition == "ko_firre")),"gene_id"])

v <- venn.diagram(list("ko_90" = ko_90, "wt_180" = wt_180), 
                       filename = NULL)
grid.newpage()
grid.draw(v)


sg_wtko <- intersect(ko_90, wt_180)
paste(unique(dat_sig[which(dat_sig$gene_id %in% sg_wtko),"gene_name"]), collapse = ";")
```

Okay, for future reference, it would be good to know what these lists of overlapping genes are. 

```{r}
# Let's get a g2s to merge in
g2s <- dat_sig %>% select(gene_id, gene_name) %>% distinct()

vp <- get.venn.partitions(list("ko_dox_90" = ko_dox_90, "ko_firre_90" = ko_firre_90)) 
vp$comparison <- paste0(names(vp)[1],"_vs_", names(vp)[2])
names(vp) <- c("group1", "group2", "set", "gene_id", "count", "comparison")
vp <- vp %>% select(-count)  
vp <- vp %>% unnest(gene_id)
vp <- merge(vp, g2s)


v <- get.venn.partitions(list("ko_dox_180" = ko_dox_180, "ko_firre_180" = ko_firre_180))
v$comparison <- paste0(names(v)[1],"_vs_", names(v)[2])
names(v) <- c("group1", "group2", "set", "gene_id", "count", "comparison")
v <- v %>% select(-count)  
v <- v %>% unnest(gene_id)
v <- merge(v, g2s)
vp <- bind_rows(vp, v)

v <- get.venn.partitions(list("wt_dox_180" = wt_dox_180, "wt_firre_180" = wt_firre_180))
v$comparison <- paste0(names(v)[1],"_vs_", names(v)[2])
names(v) <- c("group1", "group2", "set", "gene_id", "count", "comparison")
v <- v %>% select(-count)  
v <- v %>% unnest(gene_id)
v <- merge(v, g2s)
vp <- bind_rows(vp, v)

v <- get.venn.partitions(list("sg_ko_90" = sg_ko_90, "sg_wt_180" = sg_wt_180))
v$comparison <- paste0(names(v)[1],"_vs_", names(v)[2])
names(v) <- c("group1", "group2", "set", "gene_id", "count", "comparison")
v <- v %>% select(-count)  
v <- v %>% unnest(gene_id)
v <- merge(v, g2s)
vp <- bind_rows(vp, v)

v <- get.venn.partitions(list("ko_90" = ko_90, "wt_180" = wt_180))
v$comparison <- paste0(names(v)[1],"_vs_", names(v)[2])
names(v) <- c("group1", "group2", "set", "gene_id", "count", "comparison")
v <- v %>% select(-count)  
v <- v %>% unnest(gene_id)
v <- merge(v, g2s)
vp <- bind_rows(vp, v)







write.csv(vp, "deg_delta_overlaps.csv", row.names = FALSE)
```



