---
title: "Long timecourse rescue"
author: "Michael Smallegan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_notebook:
    code_folding: hide
    css: /scratch/Shares/rinn/Michael/firre_timecourse/analysis/util/style.css
    theme: simplex
    toc: yes
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(stringsAsFactors = FALSE,
        dplyr.summarise.inform = FALSE,
        tidyverse.quiet = TRUE)
library(tidyverse)
library(DESeq2)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)
library(seriation)
source("../util/_plot_theme.R")
source("../util/_util.R")
source("../01_setup/assumptions.R")
source("../util/figures.R")
# Keep track of thresholds for figure naming
thresh <- paste0("pval", pval_thresh, "_l2fc", round(l2fc_thresh, 2))
```

```{r load, include=FALSE}
load("../01_setup/results/rnaseq_data.RData")
```

For this experiment, we induced the expression of Firre with an rTTA element by adding doxycycline to mouse embryonic stem cells. We see that Firre is indeed expressed in the KO background after the addition of doxycycline. The drug does instigate some gene expression changes on its own, so we will control for the effects by using a linear model which accounts for the effect of dox.

```{r long_timecourse}
if(!file.exists("results/ko_rescue_long.RData")) {
  
  # Filter to ESC KO long timecourse
  ko_rescue_long_samples <- samples %>%
    filter(cell_type == "ESC",
           timecourse_length == "long",
           firre_ko == "KO")
  ko_rescue_long_counts <- salmon_gene_counts[,ko_rescue_long_samples$sample_id]
  
  # Check ordering
  stopifnot(all(rownames(ko_rescue_long_samples) == colnames(ko_rescue_long_counts)))
  stopifnot(all(rownames(ko_rescue_long_counts) == genes$gene_id))
  
  # DESeq2 -- controlling for doxycycline; likelihood ratio test
  ko_rescue_long_dds <- DESeqDataSetFromMatrix(countData = ko_rescue_long_counts, 
                                          colData = ko_rescue_long_samples, 
                                          design = ~ firre_induced + timepoint + timepoint*firre_induced)
  ko_rescue_long_dds <- DESeq(ko_rescue_long_dds, test = "LRT", reduced = ~ firre_induced + timepoint)
  
  # Compile results
  res_names <- resultsNames(ko_rescue_long_dds)
  dynamic_res <- res_names[grepl("firre_inducedfirre_induced.timepoint", res_names)]
  
  ko_rescue_long_lfc <- lapply(dynamic_res, function(x) {
    results(ko_rescue_long_dds, 
            name = x) %>%
      as.data.frame() %>%
      rownames_to_column("gene_id") %>% 
      merge(g2s) %>%
      mutate(result_name = x,
             timepoint = as.numeric(gsub("firre_inducedfirre_induced.timepoint", "", result_name)))
  }) %>% bind_rows()
  
  # Shrunken LFC results
  ko_rescue_long_shrnklfc <- lapply(dynamic_res, function(x) {
    lfcShrink(ko_rescue_long_dds, 
              coef = x,
              type = "apeglm") %>%
      as.data.frame() %>%
      rownames_to_column("gene_id") %>% 
      merge(g2s) %>%
      mutate(result_name = x,
             timepoint = as.numeric(gsub("firre_inducedfirre_induced.timepoint", "", result_name)))
  }) %>% bind_rows()
  
  # Calculate the maximum fold-change in any one timepoint
  ko_rescue_long_maxfc <- ko_rescue_long_shrnklfc %>%
    group_by(gene_id) %>%
    summarize(max_fc = max(abs(log2FoldChange))) 
  
  ko_rescue_long_shrnklfc <- ko_rescue_long_shrnklfc %>%
    left_join(ko_rescue_long_maxfc)
  
  save(ko_rescue_long_lfc, ko_rescue_long_shrnklfc, file = "results/ko_rescue_long_long.RData")
}

load("results/ko_rescue_long.RData")
```

### Fold changes vs zero timepoint

This is without considering the control cell line.

```{r}
if(!file.exists("results/ko_rescue_long_vs_zero.RData")) {
  
  ko_rescue_long_vszero_samples <- samples %>%
    filter(cell_type == "ESC",
           timecourse_length == "long",
           firre_ko == "KO",
           firre_induced == "firre_induced")
  ko_rescue_long_vszero_counts <- salmon_gene_counts[,ko_rescue_long_vszero_samples$sample_id]
  
  # Check ordering
  stopifnot(all(rownames(ko_rescue_long_vszero_samples) == colnames(ko_rescue_long_vszero_counts)))
  stopifnot(all(rownames(ko_rescue_long_vszero_counts) == genes$gene_id))
  
  # DESeq2 -- controlling for doxycycline; likelihood ratio test
  ko_rescue_long_vszero_dds <- DESeqDataSetFromMatrix(countData = ko_rescue_long_vszero_counts,
                                                 colData = ko_rescue_long_vszero_samples,
                                                 design = ~ timepoint)
  ko_rescue_long_vszero_dds <- DESeq(ko_rescue_long_vszero_dds)
  res_names <- resultsNames(ko_rescue_long_vszero_dds)
  
  vs_zero_res <- res_names[grepl("_vs_0", res_names)]
  ko_rescue_long_vszero_shrnklfc <- lapply(vs_zero_res, function(x) {
    lfcShrink(ko_rescue_long_vszero_dds, 
              coef = x,
              type = "apeglm") %>%
      as.data.frame() %>%
      rownames_to_column("gene_id") %>% 
      merge(g2s) %>%
      mutate(result_name = x,
             timepoint = as.numeric(gsub("timepoint_|_vs_0", "", result_name)))
  }) %>% bind_rows()
  
  save(ko_rescue_long_vszero_shrnklfc, file = "results/ko_rescue_long_vs_zero.RData")
}

load("results/ko_rescue_long_vs_zero.RData", verbose = T)
```

### Rescued genes

These are the genes that are reciprocally regulated in the KO vs WT and the Firre induced line.

```{r}
# We need the KO vs WT data
load("../04_firre_ko_vs_wt/results/wtko.RData")

wtko_res_shrnklfc <- wtko_res_shrnklfc %>%
  mutate(sig = padj <= pval_thresh & abs(log2FoldChange) > l2fc_thresh)

wtko_sig <- wtko_res_shrnklfc %>% 
  filter(sig == TRUE)

# Let's figure out which genes are rescued
rescued_genes <- wtko_sig$gene_id[wtko_sig$gene_id %in% 
                                    (ko_rescue_long_shrnklfc %>%
                                       filter(padj <= pval_thresh) %>% 
                                       pull(gene_id))]

rescue_wtko <- wtko_sig %>%
  filter(gene_id %in% rescued_genes) %>%
  dplyr::select(gene_id, gene_name, log2FoldChange) %>%
  dplyr::rename(l2fc_wtko = log2FoldChange)

rescue_ltc <- ko_rescue_long_vszero_shrnklfc %>%
  filter(gene_id %in% rescued_genes) %>%
  dplyr::select(gene_id, gene_name, log2FoldChange, timepoint) %>%
  dplyr::rename(l2fc_ltc = log2FoldChange)

rescued_df <- rescue_wtko %>%
  left_join(rescue_ltc)

rescued_df <- rescued_df %>%
  mutate(dist = abs((l2fc_wtko * -1) - l2fc_ltc))

rescued_df_closest <- rescued_df %>%
  group_by(gene_id, gene_name) %>%
  mutate(closest = min(dist) == dist,
         max_fc = max(abs(l2fc_ltc)) == abs(l2fc_ltc)) %>%
  filter(closest == TRUE)

rescued_df_closest <-  rescued_df_closest %>%
  mutate(twenty_percent = l2fc_wtko * 0.2,
         rescued = (-1*l2fc_wtko - twenty_percent)>= l2fc_ltc & l2fc_ltc>= (-1*l2fc_wtko + twenty_percent) |
           (-1*l2fc_wtko + twenty_percent)>= l2fc_ltc & l2fc_ltc>= (-1*l2fc_wtko - twenty_percent))

rescued_genes_20 <- rescued_df_closest %>% 
  filter(rescued == TRUE) %>% 
  pull(gene_name)

ggplot(rescued_df_closest, aes(x = l2fc_wtko, y = l2fc_ltc, color = rescued)) +
  geom_abline(slope = -1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = -0.8, lty = 2) +
  geom_abline(slope = -1.2, lty = 2) +
  xlim(-2,2) +
  ylim(-2,2)
ggsave(paste0("figures/long_rescue_lfc_scatter_", thresh, ".pdf"), 
       height = 3, width = 3, useDingbats = FALSE)


# Let's look at the number of genes overlapping between Firre responders and the KO vs WT
rescued_gene_ids_20 <- rescued_df_closest %>% 
  filter(rescued == TRUE) %>% 
  pull(gene_id)



combined_genes <- unique(c(rescued_genes, rescued_gene_ids_20))

mat <- cbind(
  rescued_genes = combined_genes %in% rescued_genes,
  reciprocal = combined_genes %in% rescued_gene_ids_20
)
fit2 <- euler(mat)


pdf("figures/rescued_vs_reciprocal_overlap.pdf", height = 2, width = 2)
plot(fit2, quantities = TRUE, fontsize = 8)
dev.off()

ggplot(rescued_df_closest, aes(x = l2fc_wtko, y = l2fc_ltc, color = rescued)) +
  geom_abline(slope = -1) +
  geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  geom_point(alpha = 0.7) +
  geom_abline(slope = -0.8, lty = 2) +
  geom_abline(slope = -1.2, lty = 2) +
  xlim(-2,2) +
  ylim(-2,2)

library(ggpubr)
ggplot(rescued_df_closest %>% filter(gene_name != "Firre"), aes(x = l2fc_wtko, y = l2fc_ltc)) +
  geom_point() +
    geom_hline(yintercept = 0) +
  geom_vline(xintercept = 0) +
  stat_regline_equation(label.y = 1.5) +
  stat_cor() +
  geom_smooth(method = "lm")
ggsave(paste0("figures/long_rescue_lfc_scatter_with_linear_regression", thresh, ".pdf"), 
       height = 3, width = 3, useDingbats = FALSE)

```

### Long timecourse call significant genes

We'll make the p-value cutoff based on the dox controlled model and the l2fc cutoff based on the fold change vs zero.

```{r}
ko_rescue_long_dox_sig <- ko_rescue_long_shrnklfc %>% 
  filter(padj <= pval_thresh)

ko_rescue_long_vszero_sig <- ko_rescue_long_vszero_shrnklfc %>%
  filter(gene_id %in% ko_rescue_long_dox_sig$gene_id)

ko_rescue_long_vszero_maxfc <- ko_rescue_long_vszero_sig %>%
  group_by(gene_id) %>%
  summarize(max_fc = max(abs(log2FoldChange))) 

ko_rescue_long_vszero_sig <- ko_rescue_long_vszero_sig %>%
  left_join(ko_rescue_long_vszero_maxfc)

ko_rescue_long_vszero_sig <- ko_rescue_long_vszero_sig %>%
  filter(max_fc > l2fc_thresh)

ko_rescue_long_vszero_sig <- ko_rescue_long_vszero_sig %>%
  mutate(l2fc_shrunken = log2FoldChange)

save(ko_rescue_long_vszero_sig, file = "results/ko_rescue_long_vszero_sig.RData")
```

```{r}
make_timecourse_lfc_plot(deseq_res = ko_rescue_long_vszero_sig,
                         genes_to_include = unique(ko_rescue_long_vszero_sig$gene_id),
                         y_lim = c(-6,6),
                         draw_plot = FALSE)

ggsave(paste0("figures/ko_rescue_long_lfc_", thresh, ".pdf"),
       height = 3, width = 3.5, useDingbats = FALSE)
```

```{r}
# Let's look at the number of genes overlapping between Firre responders and the KO vs WT
ko_rescue_long_sig_genes <- unique(ko_rescue_long_vszero_sig$gene_id)
wtko_sig_genes <- unique(wtko_sig$gene_id)

combined_genes <- unique(c(ko_rescue_long_sig_genes, wtko_sig_genes))

mat <- cbind(
  rescue_long = combined_genes %in% ko_rescue_long_sig_genes,
  wtko_static = combined_genes %in% wtko_sig_genes
)
fit2 <- euler(mat)

ovp <- calculate_overlap_pval(ko_rescue_long_sig_genes, wtko_sig_genes, num_measured_genes)
pdf("figures/long_timecourse_rescue_kowt_static_overlap.pdf", height = 2, width = 2)
plot(fit2, quantities = TRUE, fontsize = 8)
grid.text(paste0("P=", signif(ovp, 3)), x = 0.1, y = 0.9)
dev.off()
```



### Long timecourse Firre responders heatmap

```{r, fig.width=3, fig.height=5}
# Heatmap of fold-changes for DEGs in the rescue
# Check that there are no duplicate row names.
stopifnot(all(length(unique(ko_rescue_long_vszero_sig$gene_id)) == length(unique(ko_rescue_long_vszero_sig$gene_name))))

ko_rescue_long_lfc <- ko_rescue_long_vszero_sig %>%
  dplyr::select(gene_name, timepoint, log2FoldChange) %>%
  pivot_wider(names_from = timepoint, names_sort = TRUE, values_from = log2FoldChange) %>%
  column_to_rownames("gene_name") %>%
  as.matrix()

# Add a zero column.
ko_rescue_long_lfc <- cbind(matrix(0, nrow = nrow(ko_rescue_long_lfc), ncol = 1), ko_rescue_long_lfc)
colnames(ko_rescue_long_lfc)[[1]] <- "0"

row_ha = rowAnnotation(sig_in_ko = as.numeric(rownames(ko_rescue_long_lfc) %in% wtko_sig$gene_name),
                       rescued = as.numeric(rownames(ko_rescue_long_lfc) %in% rescued_genes_20),
                       col = list(sig_in_ko = c("1" = "black", "0" = "white"),
                                  rescued = c("1" = "black", "0" = "white")))

table(as.numeric(rownames(ko_rescue_long_lfc) %in% wtko_sig$gene_name))
table(as.numeric(rownames(ko_rescue_long_lfc) %in% rescued_genes_20))

pdf(paste0("figures/long_responders_heatmap_", thresh, ".pdf"), width = 3, height = 5)
h1 <- Heatmap(ko_rescue_long_lfc, 
        name = "l2fc",
        cluster_columns = FALSE, show_row_names = FALSE, 
        col = colorRamp2(seq(-3,3,length.out = 100), col_pal10),
        left_annotation = row_ha)
draw(h1)
dev.off()
draw(h1)
```

```{r}
load("results/ko_rescue_long_vszero_sig.RData")
# How many genes are changing in each timepoint?
deg_per_timepoint <- ko_rescue_long_vszero_sig %>%
  filter(abs(log2FoldChange) > l2fc_thresh) %>%
  mutate(direction = ifelse(log2FoldChange > 0, "up", "down")) %>%
  group_by(timepoint, direction) %>%
  summarize(ndeg = n()) %>%
  mutate(y = ifelse(direction == "up", ndeg, ndeg*-1))
deg_per_timepoint$timepoint <- factor(deg_per_timepoint$timepoint, labels = c("12", "24", "48", "96"))

ggplot(deg_per_timepoint, aes(x = timepoint, y = y, fill = direction)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0) +
  geom_text(data = deg_per_timepoint %>% filter(direction == "up"), aes(label=ndeg), position=position_dodge(width=0.9), vjust=-0.25) +
  geom_text(data = deg_per_timepoint %>% filter(direction == "down"), aes(label=ndeg), position=position_dodge(width=0.9), vjust=1.1) +
  guides(fill = FALSE) +
  ylim(-150,200)
ggsave("figures/long_rescue_ndeg_per_timepoint.pdf",
       height = 2, width = 1.75)  
```

### Firre rescued genes heatmap

We can visualize the expression levels of the genes that were significantly differentially expressed in the Firre knockout and dynamically changing in response to Firre's induction in a knockout background.

```{r rescue_heatmap, fig.width=3.5, fig.height=6}
# Retrieve the set of genes that is DE in the KO and is significantly changing
# in the KO rescue experiment. 
# Retrieve the TPM -- take the mean TPM in each condition.
rescued_genes_20 <- rescued_df_closest %>% 
  filter(rescued == TRUE) %>% 
  pull(gene_id)

wtko_samples <- samples %>% 
    filter(cell_type == "ESC", 
           timepoint == 0, 
           timecourse_length == "long")

wtko_matrix <- tpm %>% 
  filter(gene_id %in% rescued_genes_20) %>%
  pivot_longer(3:ncol(.), names_to = "sample_id", values_to = "tpm") %>%
  left_join(wtko_samples) %>%
  drop_na() %>%
  group_by(gene_name, firre_ko) %>%
  summarize(tpm = mean(tpm)) %>%
  pivot_wider(names_from = "firre_ko", 
              values_from = "tpm")

ko_rescue_long_samples <- samples %>%
    filter(cell_type == "ESC",
           timecourse_length == "long",
           firre_ko == "KO")

ko_rescue_matrix <- tpm %>% 
  filter(gene_id %in% rescued_genes_20) %>%
  pivot_longer(3:ncol(.), names_to = "sample_id", values_to = "tpm") %>%
  left_join(ko_rescue_long_samples) %>%
  drop_na() %>%
  filter(timepoint != 0) %>%
  group_by(gene_name, timepoint) %>%
  summarize(tpm = mean(tpm)) %>%
  pivot_wider(names_from = "timepoint", 
              values_from = "tpm")

comb_matrix <- wtko_matrix %>% 
  left_join(ko_rescue_matrix) %>%
  column_to_rownames("gene_name") %>%
  as.matrix()

comb_matrix <- comb_matrix[,c("KO", "720", "1440", "2880", "5760", "WT")]

comb_matrix_scaled <- t(scale(t(comb_matrix)))

p1 <- pheatmap::pheatmap(comb_matrix_scaled, cluster_cols = FALSE, cluster_rows = TRUE, show_rownames = TRUE,
                   color = col_pal10,
                   clustering_callback = olo_seriate,
                   clustering_distance_rows = "euclidean",
                   border_color = NA,
                   fontsize_row = 8,
                   breaks = seq(-1.5, 1.5, length.out = length(col_pal10)),
                   treeheight_row = 25)

ggsave(paste0("figures/long_rescue_heatmap_", thresh, ".pdf"), 
       plot = p1, width = 3.5, height = 5)
```

### Rescue PCA

```{r rescue_pca}
pca_dat <- prcomp(t(comb_matrix))
proportion_of_variance <- summary(pca_dat)$importance[2,1:2]

pca_df <- pca_dat$x[,1:2] %>%
  as.data.frame() %>%
  rownames_to_column("sample") %>%
  mutate(PC1 = PC1 * -1)

ggplot(pca_df, aes(x = PC1, y = PC2, label = sample)) +
  geom_point() +
  geom_text_repel() +
  xlab(paste0("PC1: ", round(proportion_of_variance["PC1"]*100), "%")) +
  ylab(paste0("PC2: ", round(proportion_of_variance["PC2"]*100), "%"))
ggsave(paste0("figures/long_rescue_pca_", thresh, ".pdf"), 
       width = 3.5, height = 0.9, 
       useDingbats = FALSE)
```
